\pkg{casebase}: An Alternative Framework for Survival Analysis and Comparison of Event Rates

### 2022-11-19 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\widehat{\theta}\) and its variance-covariance matrix \(V(\widehat \theta)\) which is obtained as part of the logistic regression output.
\item
  Sample \(B\) times from a multivariate normal \(N\left(\widehat{\theta}, V(\widehat\theta)\right)\).
\item
  For each sample, compute the survival function using Equation \ref{eqn:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
\end{enumerate}

@@

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g.~Breslow's estimator) or parametric estimators (e.g.~spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

\begin{Schunk}
\begin{Sinput}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
\end{Sinput}
\begin{Soutput}
#> [1] TRUE
\end{Soutput}
\begin{Sinput}
attr(pt_object, "exposure")
\end{Sinput}
\begin{Soutput}
#> [1] "ScrArm"
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
plot(pt_object, add.base.series = TRUE)
\end{Sinput}
\end{Schunk}

@@

Once a model-fit object has been returned by \texttt{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \texttt{print}, \texttt{summary}, \texttt{predict}, \texttt{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right).
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
\begin{equation}\label{eqn:CDF}
F(t) = 1 - S(t).
\end{equation}

@@

\begin{Schunk}
\begin{Sinput}
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 10)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
summary(fit) 
\end{Sinput}
\begin{Soutput}
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 5400 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ pspline(Follow.Up.Time, 
#>     df = 2) * ScrArm, data = ERSPC, ratio = 10)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -1.168  -0.486  -0.414  -0.215   3.262  
#> 
#> Coefficients:
#>                                                        Estimate Std. Error
#> (Intercept)                                              -13.81       9.98
#> pspline(Follow.Up.Time, df = 2)1                           2.66      10.96
#> pspline(Follow.Up.Time, df = 2)2                           6.43       9.73
#> pspline(Follow.Up.Time, df = 2)3                           5.57      10.10
#> pspline(Follow.Up.Time, df = 2)4                           7.27       9.90
#> pspline(Follow.Up.Time, df = 2)5                           6.54      10.10
#> pspline(Follow.Up.Time, df = 2)6                          10.82      10.03
#> pspline(Follow.Up.Time, df = 2)7                         -11.74      30.13
#> ScrArmScreening group                                      9.22      13.35
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group    -9.25      14.85
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group    -9.80      12.97
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group    -9.04      13.54
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group    -9.39      13.23
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -10.65      13.55
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group    -8.86      13.46
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -11.58      36.92
#>                                                        z value Pr(>|z|)
#> (Intercept)                                              -1.38     0.17
#> pspline(Follow.Up.Time, df = 2)1                          0.24     0.81
#> pspline(Follow.Up.Time, df = 2)2                          0.66     0.51
#> pspline(Follow.Up.Time, df = 2)3                          0.55     0.58
#> pspline(Follow.Up.Time, df = 2)4                          0.73     0.46
#> pspline(Follow.Up.Time, df = 2)5                          0.65     0.52
#> pspline(Follow.Up.Time, df = 2)6                          1.08     0.28
#> pspline(Follow.Up.Time, df = 2)7                         -0.39     0.70
#> ScrArmScreening group                                     0.69     0.49
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.62     0.53
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.76     0.45
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.67     0.50
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.71     0.48
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.79     0.43
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.66     0.51
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.31     0.75
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 3619.1  on 5939  degrees of freedom
#> Residual deviance: 3359.1  on 5924  degrees of freedom
#> AIC: 3391
#> 
#> Number of Fisher Scoring iterations: 7
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
anova(fit, test = "LRT")
\end{Sinput}
\begin{Soutput}
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                                        Df Deviance Resid. Df Resid. Dev
#> NULL                                                    5939       3619
#> pspline(Follow.Up.Time, df = 2)         7    246.6      5932       3373
#> ScrArm                                  1      5.6      5931       3367
#> pspline(Follow.Up.Time, df = 2):ScrArm  7      7.9      5924       3359
#>                                        Pr(>Chi)    
#> NULL                                               
#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
#> ScrArm                                    0.018 *  
#> pspline(Follow.Up.Time, df = 2):ScrArm    0.343    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption[Estimated hazard ratio and 95\% confidence interval for screening vs]{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
                    ScrArm : I(Follow.Up.Time - t0),
                data = ERSPC)
\end{Sinput}
\end{Schunk}

\noindent Here, we use the binary variable \texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to write the two cases of our formula above in a more compact way. We also use the function \texttt{I()}, which allows us to specify new variables within the formula using normal \texttt{R} code. This is required, as otherwise the symbol \texttt{-} would be interpreted as a formula operator. Finally, the term \texttt{ScrArm\ :\ I(Follow.Up.Time\ -\ t0)} represents the product \(Z (t - t_0)\) in the equation above. We use the operator \texttt{:} instead of \texttt{*} in order to only include the interaction term, not the main effects.

@@

\begin{Schunk}
\begin{Sinput}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
\end{Sinput}
\begin{Soutput}
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{Soutput}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption[Estimated hazard functions for control and screening groups in the ERSPC dataset]{Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
conf_ints <- confint(risk, fit)
head(conf_ints)
\end{Sinput}
\begin{Soutput}
#>   time estimate conf.low conf.high      cov_prof
#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
#> 2  0.1  1.8e-06  1.9e-07   1.2e-05 Control group
#> 3  0.2  3.9e-06  4.8e-07   2.2e-05 Control group
#> 4  0.3  6.1e-06  9.0e-07   3.2e-05 Control group
#> 5  0.4  8.7e-06  1.5e-06   4.1e-05 Control group
#> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
\end{Soutput}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption[Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands]{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption[Population-time plot for the stem-cell transplant study with both relapse and competing events]{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
                        const(Phase) + const(Source) + const(Age),
                      data = bmtcrr, cause = 1, model = "fg")

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
\end{Sinput}
\end{Schunk}

@@

For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \ref{eq:penest}. We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) (Knaus et al. 1995).\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.~The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except \texttt{sps} and \texttt{aps} were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use \texttt{fitSmoothHazard.fit}, which is a matrix interface to the \texttt{fitSmoothHazard} function. The \texttt{fitSmoothHazard} and \texttt{fitSmoothHazard.fit} functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package. The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \texttt{glmnet::cv.glmnet} and the binomial family. To \texttt{fitSmoothHazard.fit}, we supply both a matrix \texttt{y} containing the time and event variables, and a matrix \texttt{x} containing all other covariates. We apply the lasso penalty by setting \texttt{alpha\ =\ 1} and assign a \texttt{penalty.factor} (\(w_j\); cf.~Equation \ref{eq:penest}) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Compute \(\widehat{\theta}\) and its variance-covariance matrix \(V(\widehat \theta)\) which is obtained as part of the logistic regression output.
\item
  Sample \(B\) times from a multivariate normal \(N\left(\widehat{\theta}, V(\widehat\theta)\right)\).
\item
  For each sample, compute the survival function using Equation \eqref{eq:surv} below.
\item
  Use the pointwise quantiles of these survival function estimates to construct a pointwise confidence band for the survival function of interest.
\end{enumerate}

@@

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \eqref{eq:surv} and \eqref{eq:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g.~Breslow's estimator) or parametric estimators (e.g.~spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

\begin{verbatim}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
#> [1] TRUE
attr(pt_object, "exposure")
#> [1] "ScrArm"
\end{verbatim}

@@

\begin{verbatim}
plot(pt_object, add.base.series = TRUE)
\end{verbatim}

@@

Once a model-fit object has been returned by \texttt{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \texttt{print}, \texttt{summary}, \texttt{predict}, \texttt{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}
S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right). 
\label{eq:surv}
\end{equation}
The integral is computed using either the numerical or Monte-Carlo integration. The risk function (or cumulative distribution function) is then defined as
\begin{equation}
F(t) = 1 - S(t). \label{eq:CDF}
\end{equation}

@@

\begin{verbatim}
fit <- fitSmoothHazard(DeadOfPrCa ~ pspline(Follow.Up.Time, df = 2) * ScrArm, 
                       data = ERSPC, ratio = 10)
\end{verbatim}

@@

\begin{verbatim}
summary(fit) 
#> Fitting smooth hazards with case-base sampling
#> 
#> Sample size: 159893 
#> Number of events: 540 
#> Number of base moments: 5400 
#> ----
#> 
#> Call:
#> fitSmoothHazard(formula = DeadOfPrCa ~ pspline(Follow.Up.Time, 
#>     df = 2) * ScrArm, data = ERSPC, ratio = 10)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -1.168  -0.486  -0.414  -0.215   3.262  
#> 
#> Coefficients:
#>                                                        Estimate Std. Error
#> (Intercept)                                              -13.81       9.98
#> pspline(Follow.Up.Time, df = 2)1                           2.66      10.96
#> pspline(Follow.Up.Time, df = 2)2                           6.43       9.73
#> pspline(Follow.Up.Time, df = 2)3                           5.57      10.10
#> pspline(Follow.Up.Time, df = 2)4                           7.27       9.90
#> pspline(Follow.Up.Time, df = 2)5                           6.54      10.10
#> pspline(Follow.Up.Time, df = 2)6                          10.82      10.03
#> pspline(Follow.Up.Time, df = 2)7                         -11.74      30.13
#> ScrArmScreening group                                      9.22      13.35
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group    -9.25      14.85
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group    -9.80      12.97
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group    -9.04      13.54
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group    -9.39      13.23
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -10.65      13.55
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group    -8.86      13.46
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -11.58      36.92
#>                                                        z value Pr(>|z|)
#> (Intercept)                                              -1.38     0.17
#> pspline(Follow.Up.Time, df = 2)1                          0.24     0.81
#> pspline(Follow.Up.Time, df = 2)2                          0.66     0.51
#> pspline(Follow.Up.Time, df = 2)3                          0.55     0.58
#> pspline(Follow.Up.Time, df = 2)4                          0.73     0.46
#> pspline(Follow.Up.Time, df = 2)5                          0.65     0.52
#> pspline(Follow.Up.Time, df = 2)6                          1.08     0.28
#> pspline(Follow.Up.Time, df = 2)7                         -0.39     0.70
#> ScrArmScreening group                                     0.69     0.49
#> pspline(Follow.Up.Time, df = 2)1:ScrArmScreening group   -0.62     0.53
#> pspline(Follow.Up.Time, df = 2)2:ScrArmScreening group   -0.76     0.45
#> pspline(Follow.Up.Time, df = 2)3:ScrArmScreening group   -0.67     0.50
#> pspline(Follow.Up.Time, df = 2)4:ScrArmScreening group   -0.71     0.48
#> pspline(Follow.Up.Time, df = 2)5:ScrArmScreening group   -0.79     0.43
#> pspline(Follow.Up.Time, df = 2)6:ScrArmScreening group   -0.66     0.51
#> pspline(Follow.Up.Time, df = 2)7:ScrArmScreening group   -0.31     0.75
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 3619.1  on 5939  degrees of freedom
#> Residual deviance: 3359.1  on 5924  degrees of freedom
#> AIC: 3391
#> 
#> Number of Fisher Scoring iterations: 7
\end{verbatim}

@@

\begin{verbatim}
anova(fit, test = "LRT")
#> Analysis of Deviance Table
#> 
#> Model: binomial, link: logit
#> 
#> Response: DeadOfPrCa
#> 
#> Terms added sequentially (first to last)
#> 
#> 
#>                                        Df Deviance Resid. Df Resid. Dev
#> NULL                                                    5939       3619
#> pspline(Follow.Up.Time, df = 2)         7    246.6      5932       3373
#> ScrArm                                  1      5.6      5931       3367
#> pspline(Follow.Up.Time, df = 2):ScrArm  7      7.9      5924       3359
#>                                        Pr(>Chi)    
#> NULL                                               
#> pspline(Follow.Up.Time, df = 2)          <2e-16 ***
#> ScrArm                                    0.018 *  
#> pspline(Follow.Up.Time, df = 2):ScrArm    0.343    
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./interaction-ERSPC-1} \caption{Estimated hazard ratio and 95\% confidence interval for screening vs. control group as a function of time in the ERSPC dataset. Hazard ratios are estimated from fitting a parametric hazard model as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. 95\% confidence intervals are calculated using the delta method. The plot shows that the effect of screening only begins to become statistically apparent by year 7. The 25-60\% reductions seen in years 8-12 of the study suggests a much higher reduction in prostate cancer due to screening than the single overall 20\% reported in the original article.}\label{fig:interaction-ERSPC}
\end{figure}

@@

\begin{verbatim}
fitSmoothHazard(DeadOfPrCa ~ as.integer(Follow.Up.Time >= t0) :
                    ScrArm : I(Follow.Up.Time - t0),
                data = ERSPC)
\end{verbatim}

Here, we use the binary variable \texttt{as.integer(Follow.Up.Time\ \textgreater{}=\ t0)} in order to write the two cases of our formula above in a more compact way. We also use the function \texttt{I()}, which allows us to specify new variables within the formula using normal \texttt{R} code. This is required, as otherwise the symbol \texttt{-} would be interpreted as a formula operator. Finally, the term \texttt{ScrArm\ :\ I(Follow.Up.Time\ -\ t0)} represents the product \(Z (t - t_0)\) in the equation above. We use the operator \texttt{:} instead of \texttt{*} in order to only include the interaction term, not the main effects.

@@

\begin{verbatim}
plot(fit, type = "hazard",
     hazard.params = list(xvar = "Follow.Up.Time",
                          by = "ScrArm"))
#> Conditions used in construction of plot
#> ScrArm: Control group / Screening group
#> offset: 0
\end{verbatim}

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./cs1hazard-1} \caption{Estimated hazard functions for control and screening groups in the ERSPC dataset. Hazards are estimated from fitting a parametric model with casebase sampling as a function of the interaction between a cubic pspline basis (df=2) of follow-up time and treatment arm. The package vignettes provide a detailed description of how to plot hazard functions for any combination of covariates along with confidence bands.}\label{fig:cs1hazard}
\end{figure}

@@

\begin{verbatim}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))
new_time <- seq(0,14,0.1)
risk <- absoluteRisk(fit, time = new_time, newdata = new_data)
\end{verbatim}

@@

\begin{verbatim}
conf_ints <- confint(risk, fit)
head(conf_ints)
#>   time estimate conf.low conf.high      cov_prof
#> 1  0.0  0.0e+00  0.0e+00   0.0e+00 Control group
#> 2  0.1  1.8e-06  1.9e-07   1.2e-05 Control group
#> 3  0.2  3.9e-06  4.8e-07   2.2e-05 Control group
#> 4  0.3  6.1e-06  9.0e-07   3.2e-05 Control group
#> 5  0.4  8.7e-06  1.5e-06   4.1e-05 Control group
#> 6  0.5  1.2e-05  2.3e-06   5.0e-05 Control group
\end{verbatim}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./erspc-cif-conf-1} \caption{Risk function estimates for control and screening groups in the ERSPC data using case-base sampling and Kaplan-Meier, along with 95\% confidence bands.}\label{fig:erspc-cif-conf}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth,keepaspectratio=true]{./compPop-1} \caption{Population-time plot for the stem-cell transplant study with both relapse and competing events.}\label{fig:compPop}
\end{figure}

@@

\begin{verbatim}
model_cb <- fitSmoothHazard(
  Status ~ ftime + Sex + D + Phase + Source + Age,
  data = bmtcrr,
  ratio = 100,
  time = "ftime"
)
\end{verbatim}

@@

\begin{verbatim}
library(survival)
# Prepare data for coxph
bmtcrr_cox <- transform(bmtcrr, 
                        id = seq_len(nrow(bmtcrr)),
                        Status = factor(Status))

model_cox <- coxph(Surv(ftime, Status) ~ Sex + D + Phase + Source + Age,
                   data = bmtcrr_cox, id = id)
\end{verbatim}

@@

\begin{verbatim}
library(timereg)
model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
                        const(Phase) + const(Source) + const(Age),
                      data = bmtcrr, cause = 1, model = "fg")

# Estimate CI curve
risk_fg <- predict(model_fg, newdata, times = time_points)
\end{verbatim}

@@

\begin{verbatim}
# Estimate absolute risk curve
risk_cox <- survfit(model_cox, newdata = newdata)
\end{verbatim}

@@

For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by Equation \eqref{eq:penest}. We note that this is different than the semiparametric model Coxnet, which regularizes the Cox partial likelihood. To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) (Knaus et al. 1995).\footnote{The original data is available online from the Department of Biostatistics at Vanderbilt University: \url{https://biostat.app.vumc.org/wiki/Main/SupportDesc}} The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill.~The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except \texttt{sps} and \texttt{aps} were used. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opt for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use \texttt{fitSmoothHazard.fit}, which is a matrix interface to the \texttt{fitSmoothHazard} function. The \texttt{fitSmoothHazard} and \texttt{fitSmoothHazard.fit} functions sample the case and base series, calculate the required offset, and transform the data to match the expected input of the \pkg{glmnet} package. The penalized logistic regression is then fit for multiple values of the tuning parameter using the function \texttt{glmnet::cv.glmnet} and the binomial family. To \texttt{fitSmoothHazard.fit}, we supply both a matrix \texttt{y} containing the time and event variables, and a matrix \texttt{x} containing all other covariates. We apply the lasso penalty by setting \texttt{alpha\ =\ 1} and assign a \texttt{penalty.factor} (\(w_j\); cf.~Equation \eqref{eq:penest}) of 0 to the time variable to ensure it is in the selected model. We compare our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package and using the Cox family).
